Letter: Performance analysis of \\ high-resolution ice sheet simulations

### 2022-06-28 paper/perfmod.tex
ready for final draft of discussion and conclusions

\abstract{High spatial resolution numerical ice sheet models, with horizontal mesh/grid resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.  These models usually require many time steps, shorter than climate-coupling time scales, because ice thickness is updated following each velocity solution.  High-resolution performance is degraded by the stability restrictions of such explicit time-stepping.  This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, attempts to clarify numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.  The prospective performance of implicit time-stepping numerical models is assessed.  The main result, Table \ref{tab:performancemodel}, emphasizes the key role played by the performance of stress-balance and implicit-step solvers.}

@@

Numerical ice sheet (glacier) models with evolving ice geometry are now in routine use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.  It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows need finer resolution.  Whether using local mesh refinement \citep[for example]{Hoffmanetal2018} or not, resolutions of a few kilometers to one kilometer \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used at ice sheet scale.

Current-generation ice sheet models use a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, to the non-shallow and non-hydrostatic Stokes approximation.  With very few exceptions, however, current models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the just-computed velocity field.  Therefore ice thickness and surface elevation, functionally-equivalent variables for this purpose, are updated after velocity is fixed.  Interaction between the ice sheet and the surrounding climate occurs during the geometry-update operation via the mass continuity or surface kinematical equations \citep{GreveBlatter2009}.  Such climatic coupling occurs through surface mass balance, sub-shelf (basal) mass balance, and calving processes in particular.

Thus these standard schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use the velocity computed from the geometry at the previous time step \citep[for example]{Chengetal2017}.}  At least for simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}, and stability conditions of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the time-step limits of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.

However, actual \emph{implicit} time-stepping \citep{Bueler2021,LeVeque2007} should also be considered.  Here the velocity and geometry are updated by simultaneously solving mass and momentum conservation equations.  While implicit time-stepping always requires the solution of systems of equations at each step \citep{LeVeque2007}, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}.  The problem is of free-boundary type, both in map-plane (horizontal) \citep{SchoofHewitt2013} and (easily-resolved) vertical directions.

An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}, and the same reference shows how steady-state problem \citep[compare][]{JouvetBueler2012} can often be solved, demonstrating unconditional stability.  (Observe that steady-state solutions correspond to an implicit time step of infinite duration.)  The corresponding problem for the Stokes equations has not, to the author's knowledge, yet been attempted.  However, important early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} to solve for a free boundary illuminates some of the techniques and difficulties needed to make such a strategy work for a membrane-stress-resolving balance.

This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.  A simplified performance (computational work) analysis can expose the most important considerations and trade-offs.  While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping is applied, along with advanced solver techniques like multigrid \citep{Briggsetal2000}, this belief should be inspected quantitatively to the extent possible.

@@

The reader is assumed to have access to the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.  These (continuum) models are regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}.

@@

Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.  This is because ice flows dominantly downhill.  Indeed, ice sheet flow has no characteristic curves, as would \eqref{eq:masscontinuity} if it were a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.  Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of (discretized) advective and diffusive mechanisms.

Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.  Under a numerical thickness perturbation the velocity $\bu$ often responds by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational driving stress source term is along $\gradx s$.  In fact equation \eqref{eq:masscontinuity} has velocity $\bu$ which is a non-local function $\bu(H,\grad_x s)$, though a stress balance solution is required to evaluate it.  A numerical instability occurs when the ice thickness under/over-shoots its correct value because the numerical velocity response from evaluating this non-local function is too strong.

\newcommand{\nn}{\text{n}}
Neither is the mass continuity equation wholely diffusive for Stokes dynamics.  A diffusive description is valid, however, in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}.  In the isothermal case we see a revised mass continuity equation:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity and $C$ is a positive constant.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ where $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.  Typically $\nn\approx 3$.}  While equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed.  For grounded ice sheets one observes that large values of $d$ indicate locations of unstable mode growth if explicit time-steps are chosen too large.

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry-update (and velocity) solution requires $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit, coupled Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of simulation domain & km \\
$m$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step in model years & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis; $\alpha,\beta,\gamma,n$ are pure numbers.}
\label{tab:notation}
\end{table*}

@@

Table \ref{tab:notation} lists the parameters used in our simulation performance  analysis.  Our primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.  \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits, and big-O notation is only used in this limit.  If the model domain is of width $L$ then these parameters are related by
\begin{equation}
\Delta x = O\left(\frac{L}{\sqrt{m}}\right) \quad \leftrightarrow \quad m = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
\end{equation}
That is, for a given value of $L$ there are $O(\sqrt{m})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$.  However, our analysis addresses spatially-3D models with map-plane horizontal meshes.}

A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.  It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.  The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.  Ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.

The assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012,
Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.  For example, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016} solvers use 3D refinement, though with various modifications regarding vertical refinement.  This and other mesh/solver details cannot be pursued here.

Ice sheet models resolve climate interactions (coupling), especially via surface mass balance, on time scales which are dominated by an annual cycle and certain longer scales.  Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.  Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly coupling frequency, respectively.  Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated using velocity, i.e.~via solution of the mass continuity or surface kinematical equation.

@@

For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value (from equation \eqref{eq:siamasscontinuity}).  For Stokes dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but we propose that an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, or computed from modeled ice velocities, represents the \emph{optimistic} paradigm.  The corresponding \emph{pessimistic} paradigm enforces $\Delta t < O(D^{-1} \Delta x^2)$, using (otherwise diagnostic) diffusivity values computed in the SIA manner.

Explicit time-stepping with hybrid and higher-order schemes is better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.  Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}.  The optimistic time step is supported for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).

Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $1/q$ itself.  This restriction reflects the simulation purpose and not the maintenance of stability.  Indeed, for an implicit scheme the frequency of climate coupling determines the time-stepping simulation cost, according to the solution cost of coupled mass and momentum equations.

For each explicit time step the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.  A Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited.

By contrast, a multigrid method \citep{Trottenbergetal2001} will greatly reduce $\alpha$ if correctly applied.  Antarctic ice sheet results by \cite[Table 8.1]{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$.  \cite[Table 7.5]{Tuminaroetal2016} reports total algebraic-multigrid-preconditioned iterations growing as $\alpha \approx 0.05$ in a higher-order velocity solver for the Greenland ice sheet.  (Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.)
% see data/scaling.m
An algebraic-multigrid based solver by \cite{Lengetal2012} may have similar algorithmic scaling, but reported results do not constrain $\alpha$.  A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.\footnote{The \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Lengetal2012,Tumarinoetal2016} results all show good parallel scaling.}  Note that $\alpha=0$ describes an \emph{optimal} solver  \citep{Bueler2021} in the language of scientific computation.

At the easiest end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula.  Therefore, for the SIA a velocity solution requires only $O(m)$ work: $\alpha=0$.

When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, and will vary depending on solver design.  In fact, many considerations are suppressed in our simple flops-based analysis of algorithmic scaling, and actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.

Regardless of the stress balance source, an explicit time-stepping scheme then applies the mass continuity equation to update the ice thickness using $O(m)$ work.  That is, once the velocity is computed for the geometry from the previous time step, an explicit scheme replaces old thickness values by new ones using a pointwise formula.\footnote{Additional computation needed to remesh the updated geometry, a strongly design-dependent cost, is omitted here.}

@@

Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.  Recalling the scaling of $\Delta x$ with $m$, in \eqref{eq:paramrelation}, we now write time step restrictions as required numbers of time-steps per model year in order to maintain stability, as a function of horizontal resolution $\Delta x$ or degrees of freedom $m$.  In the explicit SIA and pessimistic-Stokes cases we have
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
\end{equation}
The explicit, optimistic-Stokes estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}
The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$, to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

Unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve climatic interactions.  On the other hand, the per-step expense is much greater because nontrivial equations, indeed a free-boundary problem, must be solved.  We parameterize the flops of such coupled velocity and geometry-update solutions as $O(m^{1+\beta})$, with the understanding that $\beta \ge \alpha$ for the $\alpha$ value associated to the corresponding fixed-geometry velocity solution.

To our knowledge there is only one implemented, unconditionally-stable, fully-implicit geometric-update solver, namely the SIA scheme in \cite{Bueler2016}.  For simplified dome geometry this method, based on Newton steps solved via direct linear algebra, shows $\beta=1.8$.
% see data/bueler.perf and data/scalingbueler.m
The same work shows how a steady-state ($\Delta t=\infty$) SIA solution is practical for many geometries, but in any case convergence is robust for time steps of years to centuries on kilometer-scale grids, even using realistic, thus irregular, bed topographies.

The hybrid \cite{Brinkerhoffetal2017} flowline solver is implicit but apparently subject to a time-stepping condition when the margin moves, and its algorithmic scaling is unknown.

>>>>

\abstract{Numerical ice sheet models compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.  At high spatial resolution, with horizontal mesh/grid resolutions of a few kilometers or smaller, these models usually require time steps shorter than climate-coupling time scales because these models update ice thickness after each velocity solution.  Although precise time-step conditions are often lacking, high-resolution performance is degraded by the stability restrictions of such explicit time-stepping.  This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, attempts to clarify numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.  The performance of current-generation explicit time-stepping models is assessed, and then the prospective performance of implicit schemes.  The main result, Table \ref{tab:performancemodel}, highlights the key role which will be played by the algorithmic scaling of stress-balance and implicit-step solvers.}

@@

Numerical ice sheet (glacier) models with evolving ice geometry are now in routine use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.  It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows need yet finer resolution.  Whether using local mesh refinement \citep[for example]{Fischleretal2022,Hoffmanetal2018} or not, resolutions of two or one kilometers \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used for science at ice sheet scale.

Current-generation ice sheet models use a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, up to the non-shallow and non-hydrostatic Stokes approximation.  With very few exceptions, however, current models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the just-computed velocity field.  Therefore ice thickness and surface elevation, functionally-equivalent variables for this purpose, are updated after velocity is fixed.  Note that the interaction between the ice sheet and the surrounding climate occurs during this geometry-update operation, via the mass continuity or surface kinematical equations \citep{GreveBlatter2009}.  Such climatic coupling occurs through surface mass balance, sub-shelf (basal) mass balance, and calving processes in particular.

These current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use the velocity computed from the geometry at the previous time step \citep[for example]{Chengetal2017}.}  At least for simpler, textbook partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}, and stability conditions of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the explicit time-stepping conditions of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.

However, actual \emph{implicit} time-stepping \citep{Bueler2021,LeVeque2007} should also be considered.  Here the velocity and geometry are updated simultaneously by solving coupled mass and momentum conservation equations.  While implicit time-stepping always requires the solution of systems of equations at each step, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}, namely the 3D extend of the ice (once the coupled solution converges).  The problem is of free-boundary type, both in map-plane (horizontal) \citep{SchoofHewitt2013} and (easily-resolved) vertical directions.

An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}.  The same reference shows how the steady-state problem \citep[compare][]{JouvetBueler2012} can often be solved, demonstrating unconditional stability.  (Observe that steady-state solutions correspond to an implicit time step of infinite duration.)  The corresponding problem for the Stokes equations has not, to the author's knowledge, yet been attempted.  However, important early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} to solve a free boundary problem illuminates some of the techniques and difficulties needed to make such a strategy work for a membrane-stress-resolving balance.

This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.  Our simplified performance analysis can expose the most important considerations and trade-offs.  While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping is applied, along with advanced solver techniques like multigrid \citep{Briggsetal2000}, this belief should be inspected in advance, and quantitatively to the extent possible.

@@

Reader familiarity with the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013} is assumed.  These (continuum) models are regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}.

@@

Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.  This is because ice flows dominantly downhill.  Indeed, ice sheet flow has no characteristic curves, as would \eqref{eq:masscontinuity} if it were a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.

Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of (discretized) advective and diffusive mechanisms.  Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.  Under a numerical thickness perturbation the velocity $\bu$ will often respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term is along $\gradx s$.  In fact equation \eqref{eq:masscontinuity} has velocity $\bu$ which is a non-local function $\bu(H,\grad_x s)$, though a stress balance solution is required to evaluate it.  A numerical instability occurs when the ice thickness under/over-shoots its correct value because the numerical velocity response from evaluating this non-local function is too strong for the given time step.

\newcommand{\nn}{\text{n}}
Neither is the mass continuity equation wholely diffusive for stress-complete Stokes dynamics.  A diffusive description is valid, however, in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}, and we see a revised mass continuity equation:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity, in the isothermal case, in which $C$ is a positive constant.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ where $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.  Typically $\nn\approx 3$.}  While equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed.  For grounded ice sheets one observes that large values of $d$ tend to indicate locations of unstable mode growth if explicit time-steps are chosen too large.

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry-update (and velocity) solution requires $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit, coupled Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of map-plane domain & km \\
$m$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step in model years & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis; $\alpha,\beta,\gamma,n$ are pure numbers.}
\label{tab:notation}
\end{table*}

@@

Table \ref{tab:notation} lists the parameters used in our performance analysis.  The primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.  \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits, and big-O notation is only used in this limit.  If the map-plane model domain is of width $L$ then these parameters are related by
\begin{equation}
\Delta x = O\left(\frac{L}{\sqrt{m}}\right) \quad \text{and} \quad m = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
\end{equation}
That is, for a given value of $L$ there are $O(\sqrt{m})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$.  However, our analysis addresses spatially-3D models with map-plane horizontal meshes.}

A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.  It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the mesh/grid has an \emph{a priori} bounded resolution in the vertical direction.  The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.  Such ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance computes velocity as a function of the true state variables.

Our assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012,
Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.  For example, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016} solvers use 3D refinement, though with various modifications regarding vertical refinement.  This and many other mesh/solver details cannot be pursued here.

Ice sheet models resolve climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle and certain longer scales.  Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.  Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.  Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated using ice velocity, i.e.~via solution of the mass continuity or surface kinematical equation.

@@

For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value (equation \eqref{eq:siamasscontinuity}).  For Stokes or other membrane-stress-resolving dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, represents the \emph{optimistic} paradigm.  The corresponding \emph{pessimistic} paradigm requires $\Delta t < O(D^{-1} \Delta x^2)$, using a representative diffusivity value $D$ computed in the SIA manner.

Explicit time-stepping with hybrid and higher-order schemes is somewhat better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.  Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}, but other models apparently require the user to choose a time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.  The optimistic time step is supported in theory for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)), and thus some other manner of determining the time step.

Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$.  This restriction reflects the simulation purpose, not maintenance of stability.  For an implicit scheme the frequency of climate coupling determines the total simulation cost, according to the (large) solution cost, at each time step, of coupled mass and momentum equations.

For each explicit time step of a model using a membrane-stress-resolving balance the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.  For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited.

% see data/scaling.m for results in next paragraphs

By contrast, a multigrid method \citep{Trottenbergetal2001} can greatly reduce $\alpha$.  For example, Antarctic ice sheet results by \cite[Table 8.1]{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$.\footnote{The \cite{Lengetal2012} algebraic-multigrid solver may have similar scaling, but reported results do not constrain $\alpha$.}  For higher-order stress balances, \cite{Tuminaroetal2016} report total algebraic-multigrid-preconditioned iterations following $\alpha \approx 0.05$ for the Greenland ice sheet,\footnote{See Table 7.5.  Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.} and a geometric multigrid method by \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.

Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling \citep{Bueler2021}.  At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula.  Thus an SIA velocity solution requires optimal $O(m)$, $\alpha=0$, flops.

When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, strongly depending on solver design.  Furthermore, many considerations are suppressed in a flops-based analysis of algorithmic scaling.  Actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.  For example, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results all show good parallel scaling, something we do not address.

Regardless of the stress balance, an explicit time-stepping scheme then applies the mass continuity equation to update the ice thickness using $O(m)$ work.  That is, once the velocity is computed for the previous time-step's geometry, an explicit scheme replaces old thickness values by new ones using a pointwise formula.  Any additional computation needed to remesh the updated geometry, a design-dependent cost, is omitted here.

@@

How many flops are needed to simulate one model year?  Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.  We may write time-step restrictions as required numbers of time-steps per model year to maintain stability.  That is, $\Delta t^{-1}$ is bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.  Recalling scaling \eqref{eq:paramrelation} of $\Delta x$ with $m$, in the explicit SIA and pessimistic-Stokes cases we have
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
\end{equation}
The explicit, optimistic-Stokes estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}
The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$, to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

As already explained, unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve climatic interactions.  On the other hand, the per-step expense is much greater because nontrivial coupled equations, indeed a free-boundary problem, must be solved for the velocity and geometry update simultaneously.  We parameterize the flops of such coupled solutions as $O(m^{1+\beta})$, with the understanding that $\beta \ge \alpha$, for $\alpha$ associated to a corresponding velocity-only solution, and that a large constant is present.

% regarding next paragraph, see data/bueler.perf and data/scaling.m

It appears there is only one implemented, unconditionally-stable, fully-implicit geometric-update solver, namely the SIA scheme in \cite{Bueler2016}.\footnote{The hybrid \cite{Brinkerhoffetal2017} flowline solver is partly implicit but apparently subject to a time-stepping condition when the margin moves.}  For simplified dome geometry this method, based on Newton steps solved via direct linear algebra, and on ice margin determination on a single grid, shows $\beta=1.8$.  The same work shows how a steady-state ($\Delta t=\infty$) SIA solution is practical for many geometries.  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, even using realistic, thus irregular, bed topographies.

% regarding the nMCD method proposed in [Bueler unpublished], which is expensive but multigrid, it may turn out that $\gamma$ is small
